---
title: "Fill Missing Value"
output: html_document
author: Yi Fang
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
library('rvest')
```

```{r}
url <- 'https://www.imdb.com/search/title/?title_type=feature&release_date=2020-01-01,2020-12-31&count=100'
webpage <- read_html(url)
```

If we get data using `html_nodes` directly, we will find that there are only 90 metascore, instead of 100.

```{r}
metascore_data_html <- html_nodes(webpage,'.metascore')
metascore_data <- html_text(metascore_data_html)
metascore_data <-  as.integer(metascore_data)
length(metascore_data)
metascore_data
```

According to the rvest help guide, it states that "html_node() returns a nodeset the same length as the input."
So we can first use `html_nodes` to find a subset which contains 100 matches, then use `html_node` to find the metascore of this subset. Because `html_node` will return the same length as the input (which is 100 in this case), `html_node` will fill the missing value as `NA` automatically.

For the subset, we can use any element in the webpage that contains the metascore, including but not limited to `.mode-advanced`, `.lister-item-content`, `.ratings-bar`.

```{r}
metascore_data_html <- html_node(html_nodes(webpage, '.ratings-bar'), '.metascore')
metascore_data <- html_text(metascore_data_html)
metascore_data <-  as.integer(metascore_data)
length(metascore_data)
metascore_data
```


Reference:

https://www.programmerall.com/article/4032974459/

https://rvest.tidyverse.org/reference/html_nodes.html